Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-20 appendix.tex
Responding to queries.
marginal optim (also switch back to basin hopping for stability)

We follow Section 8.6 of \cite{Semple2003-em} to state the Hadamard representation of site split generating probabilities.
%das: @EM thoughts on including all of this in full generality? should we just reference using the sentence above and be done with it?
For each edge $e$ define the edge fidelity for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
We follow with an expository example computing these probabilities and likelihoods.

@@

@@

\begin{proof}
In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
Since
$$
\alpha_1 + \alpha_2 = \frac{1}{2}\beta
$$
and
$$
\alpha_1 - \alpha_2 = xy^2,
$$
simplifying \eqref{eq:restricted-bl-result} further yields
\begin{equation}
\label{eq:general-bl-result}
-\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
\end{equation}
Given the bounds
$$
\gamma_{L}(x, y) \le \gamma(\hat{x}', \hat{y}') \le \gamma_{U}(x, y)
$$
and
$$
\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}'),
$$
if
$$
-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
$$
then \eqref{general-bl-result} holds, and this is an inequality involving only $x$ and $y$.
To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
\end{proof}

>>>>

We follow Section 8.6 of \cite{Semple2003-em} to state the Hadamard representation of site split generating probabilities.
%das: @EM thoughts on including all of this in full generality? should we just reference using the sentence above and be done with it?
%EM You already have it in here, and so I don't see a reason to yank it now.
For each edge $e$ define the edge fidelity for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
We follow with an expository example computing these probabilities and likelihoods.

@@

\subsection*{Theorems and proofs}

@@

\begin{proof}
In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
Since
$$
\alpha_1 + \alpha_2 = \frac{1}{2}\beta
$$
and
$$
\alpha_1 - \alpha_2 = xy^2,
$$
simplifying \eqref{eq:restricted-bl-result} further yields
\begin{equation}
\label{eq:general-bl-result}
-\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
\end{equation}
Given the bounds
$$
\gamma_{L}(x, y) \le \gamma(\hat{x}', \hat{y}') \le \gamma_{U}(x, y)
$$
and
$$
\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}'),
$$
if
$$
-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
$$
then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x$ and $y$.
To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
\end{proof}

%EM Wanna add a short section about numerical optimization? You've seen that it's not entirely trivial.

### 2017-11-20 joint_inf.tex
marginal optim (also switch back to basin hopping for stability)

We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
%EM Do you know that "topology" means tree without branch lengths? Therefore I think "tree" is more appropriate in the following sentence and elsewhere.
In the classical configuration of each of these topologies, the interior branch length parameters are equal to the bottom two parameters.
We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
%EM Wouldn't it be better to just parameterize these with x,y,w without primes? Also, it looks super pro if we can use tex fonts in this figure too. I love the mathjax fonts, which I think you can get https://packages.ubuntu.com/trusty/fonts/fonts-mathjax . LMK if you run into trouble. The font I use is "MathJax_Math".
%das: I think I'm either going to go with something other than the primes, like z_1, z_2, z_3, or replace x and y with greek letters or whatever, but it'll be a little bit of a tedious edit.
%EM You mean globally in the whole document? Let's kick around ideas if so.
%das: I think we just need the true generating parameters---x^* and y^*---and the estimands---x', y', w'. I've been bad about using x and y instead of x^* and y^* for function arguments because it's cleaner, and I think there is inconsistency in using the primes and not. How about Greek for unknown (truth) and Latin for estimands? \alpha and \beta for generating parameters and x, y, z for estimands? Would have to change some other notation but shouldn't be too hard I hope.
%EM Wow, this is going to be some work. Sounds like a good plan, but I'd prefer for a system in which it's easy to remember which greek goes with which latin. could we do a, b, g and alpha, beta, gamma or something? Let's keep discussing before making the big change.
\includegraphics[width=.95\textwidth]{farris_blank}
\caption[short]{Farris topology $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein topology $\tau_2$}
\end{subfigure}
\caption{Two four-taxon topologies}
\label{fig:farris-fels-top}
\end{figure}

@@

The estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
%das: TODO double check this; do the same empirical plot for the restricted case
This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001.svg --in-pkl-name figures/w-hat-empirical-001.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-001}
\caption{
    %das: note to update caption
    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using differential evolution \cite{Storn1997}.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
%EM Re LBA I was only thinking of putting a little sentence here that says that thus this is _not_ long branch attraction.
%EM I don't know what "similar character" means here. I guess you are talking about his Figure 2?
Looking at the figure in \cite{Felsenstein1978-rr}, the region in Fig.~\ref{fig:inconsistency-farris} has a similar character, albeit a smaller region.
Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.

>>>>

We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
%EM Do you know that "topology" means tree without branch lengths? Therefore I think "tree" is more appropriate in the following sentence and elsewhere.
In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
%EM Wouldn't it be better to just parameterize these with x,y,w without primes? Also, it looks super pro if we can use tex fonts in this figure too. I love the mathjax fonts, which I think you can get https://packages.ubuntu.com/trusty/fonts/fonts-mathjax . LMK if you run into trouble. The font I use is "MathJax_Math".
%das: I think I'm either going to go with something other than the primes, like z_1, z_2, z_3, or replace x and y with greek letters or whatever, but it'll be a little bit of a tedious edit.
%EM You mean globally in the whole document? Let's kick around ideas if so.
%das: I think we just need the true generating parameters---x^* and y^*---and the estimands---x', y', w'. I've been bad about using x and y instead of x^* and y^* for function arguments because it's cleaner, and I think there is inconsistency in using the primes and not. How about Greek for unknown (truth) and Latin for estimands? \alpha and \beta for generating parameters and x, y, z for estimands? Would have to change some other notation but shouldn't be too hard I hope.
%EM Wow, this is going to be some work. Sounds like a good plan, but I'd prefer for a system in which it's easy to remember which greek goes with which latin. could we do a, b, g and alpha, beta, gamma or something? Let's keep discussing before making the big change.
\includegraphics[width=.95\textwidth]{farris_blank}
\caption[short]{Farris tree $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein tree $\tau_2$}
\end{subfigure}
\caption{Two four-taxon trees}
\label{fig:farris-fels-top}
\end{figure}

@@

The estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001-marginal.svg --out-pkl-name figures/w-hat-empirical-001-marginal.pkl --n-jobs 16 --marginal
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001-marginal.svg --in-pkl-name figures/w-hat-empirical-001-marginal.pkl
% if fit already
% !! TODO: update with finer-grained plot; this is a placeholder as the global optimization runs...
\includegraphics[width=\textwidth]{w-hat-empirical-001-marginal}
\caption[short]{Marginal inference optimizing \eqref{eq:marginal_likelihood}}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --out-pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001.svg --in-pkl-name figures/w-hat-empirical-001.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-001}
\caption[short]{Joint inference optimizing \eqref{eq:profile_likelihood}}
\end{subfigure}
\caption{
    %das: note to update caption
    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using differential evolution \cite{Storn1997}.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
%EM Re LBA I was only thinking of putting a little sentence here that says that thus this is _not_ long branch attraction.
%EM I don't know what "similar character" means here. I guess you are talking about his Figure 2?
Looking at the figure in \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
